---
title: "Predictive Practical"
author: "Daniel D. Mondal"
date: "2026-05-08"
output: word_document
---

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
```

## 1. Executive Summary

This document applies a comprehensive suite of statistical modeling techniques to predict student exam_score based on the student.csv dataset. The techniques explored include Ordinary Least Squares (OLS), Subset Selection, Regularization (Ridge, LASSO, Elastic Net) and Non-Parametric modeling (KNN).

## 2. Environment Setup & Data Preparation

First, we load the required packages, import the dataset, and prepare our training and testing environments. Because the dataset is large ($N = 80,000$), we implement an 80/20 train-test split.

```{r data_prep}
# Load required libraries
library(tidyverse)
library(glmnet)
library(pls)
library(caret)
library(car)
library(knitr)

# Set global seed for exact reproducibility
set.seed(2026)

# Load the dataset
df = read.csv("student.csv", stringsAsFactors = TRUE)

# Data Cleaning
# Remove 'student_id' as it has no predictive value and drop missing targets
df_model = df %>% 
  select(-student_id) %>%
  drop_na(exam_score) 

# Create Train / Test Split (80% / 20%)
train_index = createDataPartition(df_model$exam_score, p = 0.8, list = FALSE)
train_data = df_model[train_index, ]
test_data  = df_model[-train_index, ]

cat("Training Set Rows:", nrow(train_data), "\n")
cat("Testing Set Rows:", nrow(test_data), "\n")
```

## 2.5 Exploratory Data Analysis (EDA)

Before building predictive models, it is crucial to understand the geometry of our data, the distribution of our target variable, and the relationships between predictors. 

```{r eda_setup, message=FALSE, warning=FALSE}
# Load additional libraries for visualization
library(ggplot2)
library(corrplot)
library(gridExtra) # For arranging plots side-by-side
```

### 2.5.1 Target Variable Distribution (`exam_score`)

We begin by examining the distribution of our target variable. Understanding its shape helps us determine if transformations are needed to satisfy the normality assumptions of linear regression.

```{r eda_target_dist, fig.cap="Distribution of Final Exam Scores"}
ggplot(train_data, aes(x = exam_score)) +
  geom_histogram(aes(y = ..density..), bins = 40, fill = "steelblue", color = "white") +
  geom_density(color = "darkred", size = 1) +
  theme_minimal() +
  labs(title = "Distribution of Exam Scores",
       subtitle = "Density curve overlaid on histogram",
       x = "Exam Score",
       y = "Density")
```

*Observation:* The scores range from the mid-30s to 100, with a heavy left skew (a ceiling effect near 100). The mean is extremely high (~89), indicating a generally high-performing student body.

### 2.5.2 Correlation Matrix (Numeric Predictors)

To identify severe multicollinearity and find strong linear predictors, we generate a correlation matrix for all numeric variables.

```{r eda_correlation, fig.width=10, fig.height=8}
# Select only numeric columns
numeric_data = train_data %>% select(where(is.numeric))

# Calculate Pearson correlation matrix
cor_matrix = cor(numeric_data, use = "complete.obs")

# Plot using corrplot
corrplot(cor_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, addCoef.col = "black", 
         number.cex = 0.6, diag = FALSE, 
         title = "Correlation Matrix of Numeric Variables",
         mar = c(0,0,1,0))
```

*Key Takeaways:*

1. **Strongest Predictor:** `previous_gpa` has an extraordinarily high correlation with `exam_score` ($r \approx 0.93$). 

2. **Moderate Predictors:** `motivation_level` ($0.25$) and `study_hours_per_day` ($0.24$) show moderate positive correlations.

3. **Negative Predictors:** `exam_anxiety_score` ($-0.24$) and `stress_level` ($-0.12$) drag scores down.

4. **Multicollinearity:** There don't appear to be dangerously high correlations *between* the predictors themselves, which is good news for our baseline OLS model.

### 2.5.3 Bivariate Analysis: Strongest Numeric Predictors

Let's visualize the relationship between our two strongest numeric predictors and the target variable using scatter plots with a smoothing line.

```{r eda_scatterplots, fig.width=10, fig.height=5}
p1 = ggplot(train_data, aes(x = previous_gpa, y = exam_score)) +
  geom_point(alpha = 0.1, color = "darkgreen") + # Low alpha due to 64k rows
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  theme_minimal() +
  labs(title = "Previous GPA vs Exam Score", x = "Previous GPA", y = "Exam Score")

p2 = ggplot(train_data, aes(x = study_hours_per_day, y = exam_score)) +
  geom_point(alpha = 0.1, color = "navy") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  theme_minimal() +
  labs(title = "Study Hours vs Exam Score", x = "Study Hours per Day", y = "Exam Score")

grid.arrange(p1, p2, ncol = 2)
```

*Observation:* The relationship with `previous_gpa` is almost perfectly linear, confirming why it dominated the correlation matrix. `study_hours_per_day` also shows a clear positive, linear trend. This linearity suggests that parametric models (like OLS, Ridge, and LASSO) will perform very well.

### 2.5.4 Bivariate Analysis: Categorical Predictors

Finally, let's explore if certain categorical factors influence exam scores using boxplots.

```{r eda_boxplots, fig.width=12, fig.height=5}
p3 = ggplot(train_data, aes(x = major, y = exam_score, fill = major)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Exam Score by Major", x = "Major", y = "Exam Score")

p4 = ggplot(train_data, aes(x = internet_quality, y = exam_score, fill = internet_quality)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Exam Score by Internet Quality", x = "Internet Quality", y = "Exam Score")

grid.arrange(p3, p4, ncol = 2)
```

*Observation:* The median scores across different majors appear very similar, suggesting `major` might not be a massive driver of variance. However, `internet_quality` might show slight variations, emphasizing the need for our models to evaluate categorical dummy variables carefully.

---

## 3. Ordinary Least Squares (OLS) Regression

We establish our baseline by fitting a standard multiple linear regression using all available predictors.

```{r ols}
ols_model = lm(exam_score ~ ., data = train_data)

# Multicollinearity check (Variance Inflation Factor)
vif_values = vif(ols_model)[,1:2]
vif_values
# Evaluation
ols_preds = predict(ols_model, newdata = test_data)
ols_rmse = RMSE(ols_preds, test_data$exam_score)
cat("OLS Test RMSE:", round(ols_rmse, 4), "\n")
```

---

## 4. Subset Selection (Stepwise)

To reduce model complexity, we use backward stepwise selection based on the Akaike Information Criterion (AIC).
```{r stepwise, results='hide'}
# trace = 0 hides the iterative output logs
stepwise_model = step(ols_model, direction = "backward", trace = 0)
```

```{r stepwise_eval}
stepwise_preds = predict(stepwise_model, newdata = test_data)
stepwise_rmse = RMSE(stepwise_preds, test_data$exam_score)
cat("Stepwise Test RMSE:", round(stepwise_rmse, 4), "\n")
```

---

## 5. Regularization (Shrinkage Methods)

The `glmnet` package requires the input data to be formatted as matrices. This process automatically handles the conversion of categorical variables into dummy variables.

```{r matrix_prep}
# Prepare design matrices (removing the intercept column)
x_train = model.matrix(exam_score ~ ., train_data)[, -1]
y_train = train_data$exam_score
x_test  = model.matrix(exam_score ~ ., test_data)[, -1]
y_test  = test_data$exam_score
```

### 5.1 Ridge Regression ($L_2$ Penalty)

Ridge regression shrinks coefficients to manage multicollinearity without entirely removing any variables.

```{r ridge}
ridge_cv = cv.glmnet(x_train, y_train, alpha = 0)
best_lambda_ridge = ridge_cv$lambda.min

ridge_model = glmnet(x_train, y_train, alpha = 0, lambda = best_lambda_ridge)
ridge_preds = predict(ridge_model, s = best_lambda_ridge, newx = x_test)
ridge_rmse = RMSE(ridge_preds, y_test)
cat("Ridge Test RMSE:", round(ridge_rmse, 4), "\n")
```

### 5.2 LASSO Regression ($L_1$ Penalty)

LASSO acts as an automated feature selector by forcing the coefficients of irrelevant variables exactly to zero.
```{r lasso}
lasso_cv = cv.glmnet(x_train, y_train, alpha = 1)
best_lambda_lasso = lasso_cv$lambda.min

lasso_model = glmnet(x_train, y_train, alpha = 1, lambda = best_lambda_lasso)
lasso_preds = predict(lasso_model, s = best_lambda_lasso, newx = x_test)
lasso_rmse = RMSE(lasso_preds, y_test)
cat("LASSO Test RMSE:", round(lasso_rmse, 4), "\n")
```

### 5.3 Elastic Net (Hybrid)

Elastic net combines both $L_1$ and $L_2$ penalties.

```{r enet}
enet_cv = cv.glmnet(x_train, y_train, alpha = 0.5)
best_lambda_enet = enet_cv$lambda.min

enet_model = glmnet(x_train, y_train, alpha = 0.5, lambda = best_lambda_enet)
enet_preds = predict(enet_model, s = best_lambda_enet, newx = x_test)
enet_rmse = RMSE(enet_preds, y_test)
cat("Elastic Net Test RMSE:", round(enet_rmse, 4), "\n")
```

---

## 6. K-Nearest Neighbours (KNN) Regression

Given the computational intensity of calculating distance matrices for $n = 64,000$ training rows, we utilize a 5% stratified sample to train the KNN model. Center and scaling are applied to ensure distance metrics evaluate features fairly.
```{r knn}
# Create a 5% stratified sample for KNN training
set.seed(2026)
knn_sample_idx = createDataPartition(train_data$exam_score, p = 0.05, list = FALSE)
knn_train_data = train_data[knn_sample_idx, ]

# Train KNN with cross-validation
knn_model = train(
  exam_score ~ ., 
  data = knn_train_data, 
  method = "knn",
  preProcess = c("center", "scale"),
  tuneGrid = expand.grid(k = c(5, 10, 15)), 
  trControl = trainControl(method = "cv", number = 5)
)

knn_preds = predict(knn_model, newdata = test_data)
knn_rmse = RMSE(knn_preds, test_data$exam_score)
cat("KNN Test RMSE:", round(knn_rmse, 4), "\n")
```

---

## 8. Final Performance Comparison

We compile the Root Mean Squared Errors (RMSE) of all applied techniques to determine the optimal predictive model. Lower RMSE indicates superior predictive accuracy.
```{r comparison}
# Compile Results
results = data.frame(
  Model = c("OLS Baseline", "Stepwise", "Ridge (L2)", "LASSO (L1)", 
            "Elastic Net","KNN (Sampled)"),
  Test_RMSE = c(ols_rmse, stepwise_rmse, ridge_rmse, lasso_rmse, 
                enet_rmse, knn_rmse)
)

# Sort and display the Comparison
final_compare = results %>% 
  arrange(Test_RMSE) %>%
  mutate(Test_RMSE = round(Test_RMSE, 4))

kable(final_compare, caption = "Model Performance Comparison (Sorted by Test RMSE)")
```